Ecosystem size and complexity dictate riverine biodiversity

### 2021-01-26 manuscript.Rmd
Update manuscript.Rmd

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  word_document: default
  html_notebook: default
abstract:
  The species-area relationship (the SAR), a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and formed the basis of area-based conservation strategies. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity), and its role in scaling biodiversity is largely unknown. Here, we hypothesized that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity (area x habitat heterogeneity) that regulate watershed-scale species richness. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two biogeographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the conflicts between biodiversity conservation and societal demands.

@@

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches or local communities[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a] (**Figure 1**) where two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see **Methods** for model details).

@@

Influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**Figures 2 and S5-11**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity[@hubbellUnifiedNeutralTheory2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$; **Figures S13-18**). Under this scenario, branching may not contribute to the ecosystem's overall habitat diversity because local environmental noise masks environmental differences between branches. Therefore, this landscape scenario decouples the intimate relationship between branching structure and total habitat diversity, thereby eliminating the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams with diverse geological and/or climatic backgrounds show distinct environmental conditions, including water temperature, water chemistry, substrate, and flow/sediment regimes[@altermattDiversityRiverineMetacommunities2013; @bendaNetworkDynamicsHypothesis2004; @mooreEmergentStabilityLarge2015; @nakamuraDisturbanceRegimesStream2000; @petersonModellingDendriticEcological2013; @teruiMetapopulationStabilityBranching2018]. Instead, it may be more relevant to severely altered landscapes where human disturbance compromises the environmental distinctiveness of branches through, for example, flow regulations by dams[@poffHomogenizationRegionalRiver2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing to the crucial role of habitat diversity produced by branching structure.

@@

The emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation[@poffHomogenizationRegionalRiver2007], landscape modification[@finlayStreamSizeHuman2011; @moritaEffectsHabitatFragmentation2002], and stream burial[@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research[@koningNetworkGrassrootsReserves2020; @stollSmallImpoverishedRegional2013; @sundermannRiverRestorationSuccess2011]. While large reserves or restoration sites are undoubtedly important[@koningNetworkGrassrootsReserves2020], our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin[@koningNetworkGrassrootsReserves2020].

@@

### Simulation
We set four landscape and eight ecological scenarios, resulting in a total of `r n_scenario` sets of parameter combinations. For landscape scenarios, we changed values of environmental variation at headwaters ($\sigma_h$) and the degree of local environmental noise ($\sigma_l$) separately to create contrasting patterns of habitat heterogeneity. Specifically, we used two values (0.01 and 1) for each of the SD parameters. When $\sigma_h > \sigma_l$, branching exhibits important effects on habitat heterogeneity in the ecosystem because headwaters are the primary source of environmental variation. When $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity in the ecosystem.

@@

Climate (air temperature and precipitation) and land use (fraction of forest and dam density) differed clearly between the regions. Therefore, we used residuals of the simple linear regressions between the climate or land use variable (response variable) and region (explanatory variable). The residuals represent deviations from the regional averages; positive values indicate values greater than the regional average while negative values indicating values smaller than the regional average. Therefore,

>>>>

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  word_document: default
  html_notebook: default
abstract:
  The species-area relationship (the SAR), a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and formed the basis of area-based conservation strategies. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity), and its role in scaling biodiversity is largely unknown. Here, we hypothesized that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity (area x habitat heterogeneity), two factors that regulate watershed-scale species richness. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two biogeographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the conflicts between biodiversity conservation and societal demands.

@@

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches where local communities inhabit[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a] (**Figure 1**). Two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see **Methods** for model details).

@@

Influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**Figures 2 and S5-11**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity[@hubbellUnifiedNeutralTheory2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$; **Figures S13-18**). Under this scenario, branching may not contribute to the ecosystem's habitat heterogeneity because local environmental noise masks environmental differences between tributaries. Therefore, this landscape scenario decouples the intimate relationship between branching structure and habitat heterogeneity, thereby eliminating the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams with diverse geological and/or climatic backgrounds show distinct environmental conditions, including water temperature, water chemistry, substrate, and flow/sediment regimes[@altermattDiversityRiverineMetacommunities2013; @bendaNetworkDynamicsHypothesis2004; @mooreEmergentStabilityLarge2015; @nakamuraDisturbanceRegimesStream2000; @petersonModellingDendriticEcological2013; @teruiMetapopulationStabilityBranching2018]. Instead, it may be more relevant to severely altered landscapes where human disturbance compromises the environmental distinctiveness of branches through, for example, flow regulations by dams[@poffHomogenizationRegionalRiver2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing to the crucial role of habitat diversity produced by branching structure.

@@

The emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation[@poffHomogenizationRegionalRiver2007], landscape modification[@finlayStreamSizeHuman2011; @moritaEffectsHabitatFragmentation2002], and stream burial[@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research[@koningNetworkGrassrootsReserves2020; @stollSmallImpoverishedRegional2013; @sundermannRiverRestorationSuccess2011]. While large reserves or restoration sites are undoubtedly important[@koningNetworkGrassrootsReserves2020], our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin[@koningNetworkGrassrootsReserves2020]. Hence, our findings offer the ecological foundation for the complexity-based conservation strategies in rivers.

@@

### Simulation
Based on the results of extensive sensitivity analysis (**Supplementary Information**), we set four landscape and eight ecological scenarios, resulting in a total of `r n_scenario` sets of parameter combinations. For landscape scenarios, we changed values of environmental variation at headwaters ($\sigma_h$) and the degree of local environmental noise ($\sigma_l$) separately to create contrasting patterns of habitat heterogeneity. Specifically, we used two values (0.01 and 1) for each of the SD parameters. When $\sigma_h > \sigma_l$, branching exhibits important effects on habitat heterogeneity in the ecosystem because headwaters are the primary source of environmental variation. When $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity in the ecosystem.

@@

Climate (air temperature and precipitation) and land use (fraction of forest and dam density) differed clearly between the regions. Therefore, we used residuals of the simple linear regressions between the climate or land use variable (response variable) and region (explanatory variable). The residuals represent deviations from the regional averages; positive values indicate values greater than the regional average while negative values indicating values smaller than the regional average.
